Online resolving assembly graph by ONT long reads data

### 2019-06-24 paper.tex
move maths to supp

% For pseudocode
\usepackage[linesnumbered,boxed,ruled,vlined]{algorithm2e}
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
\SetCommentSty{mycommfont}
% For code embedding (Bash, Java...)
\usepackage{listings}
\lstset{basicstyle=\ttfamily,
  showstringspaces=false,
  commentstyle=\color{red},
  keywordstyle=\color{blue}
}

@@

Basically the work flow of \npgraph{} consists of 3 main modules: abundance binning, graph resolving and assembly output. 
The first step is to bin the contigs into different groups of \emph{population} based on their coverage. 
Each population would include contigs of the similar abundance in the final assembly sequences, \EG{} chromosome , plasmids, or even particular species genome in a metagenomics community.
The binning phase would assist to differentiate between repetitive contigs and unique ones. By using this information, in combination with path inducing from long reads, the assembly graph is then traversed and resolved in real-time. Finally, the graph is subjected to the last attempt of resolving and cleaning, as well as output the final results. The whole process can be managed by using either command-line interface or GUI.

\subsection*{Initial binning contigs into groups of abundances}
Each contig is represented as a node in the assembly graph and the edge between two nodes indicate their overlap (link) properties.
This step is to cluster the longest nodes (longer than $10$Kbp) into different sets, namely \emph{core} groups, based on their coverage values.

DBSCAN clustering algorithm \cite{Ester96adensity-based} is applied for this task.
The rationale is to approximate a coverage value of a significant contig (which consists of more than 10,000 \emph{k-mers}) to be a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is a distance function based on Kullback-Leibner divergence \cite{Kullback1951information}, or relative entropy, of two Poisson distributions.

Assume there are 2 Poisson distributions $P_1$ and $P_2$ with density functions $$p_1(x,\lambda_1)=\frac{e^{-\lambda_1}\lambda_1^x}{\Gamma(x+1)}$$ and $$p_2(x,\lambda_2)=\frac{e^{-\lambda_2}\lambda_2^x}{\Gamma(x+1)}$$ 
The Kullback-Leibner divergence from $P_2$ to $P_1$ is defined as:
$$D_{KL}(P_1||P_2)=\int_{-\infty}^{\infty} p_1(x)\log{\frac{p_1(x)}{p_2(x)}} dx$$
or in other words,  it is the expectation of the logarithmic difference between the probabilities $P_1$ and $P_2$, where the expectation is taken with regard to $P_1$.

The log ratio of the density functions is
$$\log{\frac{p_1(x)}{p_2(x)}}=x\log{\frac{\lambda_1}{\lambda_2}}+\lambda_2-\lambda_1$$
take expectation of this expression with regard to $P_1$ with mean $\lambda_1$ we have
$$D_{KL}(P_1||P_2)=\lambda_1\log{\frac{\lambda_1}{\lambda_2}}+\lambda_2-\lambda_1$$

The metric we used is the distance defined as
$$D(P_1,P_2)=\frac{D_{KL}(P_1||P_2)+D_{KL}(P_2||P_1)}{2}=\frac{1}{2}(\lambda_1-\lambda_2)(\log{\lambda_1}-\log{\lambda_2})$$

\subsection*{Coverage re-estimation}
Due to the possible divergence of sequencing coverage relative to the real abundance of sequences, especially the short ones, an optimization step is implemented to alleviate this issue. The coverage measure of nodes (which represent contigs) are spread throughout the graph via edges that connect them for comparison and calibration. 
The assignment of coverage to edges is also helpful to identify multiplicity in later step.

The re-estimation is basically carried out by following two steps.
\begin{itemize}
\item[1.] From nodes coverage, estimate edges' value by quadratic unconstrained optimization of the least-square function:
$$\frac{1}{2}\sum_{i}{l_i((\sum{e^{+}_{i}}-c_i)^2+(\sum{e^{-}_{i}}-c_i)^2}$$
where $l_i$ and $c_i$ is the length and coverage of a node $i$ in the graph;

$\sum{e^{+}_{i}}$ and $\sum{e^{-}_{i}}$ indicates sum of the values of incoming and outgoing edges from $i$ respectively. 
The above function and be rewritten as:
$$f(x)=\frac{1}{2}x^TQx + b^Tx + r$$
and then being minimized by using Newton or gradient method.
\item[2.] Update nodes' coverage based on itself and its neighbor edges' measures.
\end{itemize}
The calibration is iterative until no further improvements are made or a threshold loop count is reached.
\subsection*{Multiplicity estimation}
Based on the coverage values of all the edges and the graph's topology, we induce the copy numbers of every significant nodes (long contigs) in the final paths.
For each node, this could be done by investigating its adjacent edges and answering the questions of how many times it should be visited, from which abundance groups.
Multiplicities of insignificant nodes (of sequences with length less than $1,000$ bp) can be estimated in the same way but usually with less confident due to more complicated connections and greater variances of coverage values. 
For that reason, they are only used as augmented information to calculate candidate paths' score in the next step.

@@

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Bridge $B=(\overrightarrow{v_1}, \overrightarrow{v_2})$ with two ending unique bidirected nodes $\overrightarrow{v_1}, \overrightarrow{v_2}$}
\KwOut{Set of candidate paths $P$ connecting $B$}
\Begin{
$d$:=$B.length()$ \tcp*{length of the bridge or the distance between 2 ending nodes}
$M$:=$\mathtt{shortestTree}(\overrightarrow{v_2},d)$ \tcp*{build shortest tree from $\overrightarrow{v_2}$ with range $d$}
\If{$M.contain(\overrightarrow{v_1})$}{
    $S$:=new $Stack()$ \tcp*{stack of sets of edges to traverse}
    $edgesSet$:=$getEdges(\overrightarrow{v_1})$ \tcp*{get all bidirected edges going from $\overrightarrow{v_1}$}
    $S.push(edgesSet)$\;
    $p$:=new $Path(\overrightarrow{v_1})$ \tcp*{init a path that has $\overrightarrow{v_1}$ as root}
    \While{true}{
        $edgesSet$:=$S.peek()$\;
        \If{$edgesSet.isEmpty()$}{
            \If{$p.size() \leq 1$}{
                $\mathbf{break}$ \tcp*{stop the loop when there is no more edge to discover}
            }
        $S.pop()$\;
        $d$+=$p.peekNode.length()+p.popEdge().length()$\; 
        }
        \Else{
            $curEdge \coloneqq edgesSet.remove()$\;
            $\overrightarrow{v}$:=$curEdge.getOpposite(p.peekNode())$\;
            $S.push(getEdges(\overrightarrow{v}).includedIn(M))$\;
            $p.add(curEdge)$\;
            \If{reach $\overrightarrow{v_2}$ with reasonable $d$}{
                $P.add(p)$\;
            }
            $d$-=$\overrightarrow{v}.length()+curEdge.length()$\;
        }
    }
}

\Return{$P$}
}
\caption{Pseudo-code for finding paths connecting a bridge with 2 ends.}
\label{algo:findpath}
\end{algorithm}
% formal mathematical definition & DFS algorithm goes here...
In \npscarf{} with assembly graph, the path finding algorithm is the original DFS (depth first search) which becomes computationally expensive when the traversing depth increases. For \npgraph{}, we implement a modified stack-based version utilizing Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space.

Algorithm~\ref{algo:findpath} demonstrates the path finding module in general.
In which, function

$\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n$

from line 3 of the algorithm's pseudo code builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance regarding nanopore read error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

Basically, the algorithm keeps track of a stack that contains sets of candidate edges to discover. During the traversal, a variable $d$ is updated as an estimation for the distance to the target. A hit is reported if the target node is reached with a reasonable distance \IE{} close to zero, within a given tolerance (line 21). 
A threshold for the traversing depth is set (150) to ignore too complicated and time-consuming path searching.

It is worth to mention that the $length()$ functions for node and edge are totally different. While the former returns the length of the sequence represented by the node, \IE{} contig from short-read assembly, the latter is usually negative because an edge models a link between two nodes, which is normally an overlap (except for composite edges). For example, in a \emph{k-mers} DBG-derived assembly graph, the value of an edge is $-k$.

>>>>

@@

Basically the work flow of \npgraph{} consists of 3 main modules: (1) assembly graph loading and preprocessing; (2) graph resolving; (3) postprocessing and reporting. 
The first step is to load the graph of Illumina contigs and retrieve their metadata which are helpful for the next steps, \EG binning and multiplicity estimation.
By using this information, in combination with path inducing from long reads, the assembly graph is then traversed and resolved in real-time.
Finally, the graph is subjected to the last attempt of resolving and cleaning, as well as output the final results. The whole process can be managed by using either command-line interface or GUI.
Among three phases, only the first one must be performed prior to the MinION sequencing process in a streaming setup.
\subsection*{Contigs binning}
Each population is expected to include contigs of the similar abundance in the final assembly sequences, \EG{} chromosome , plasmids, or even particular species genome in a metagenomics community. In addition, the binning phase would assist to differentiate between repetitive contigs and unique ones.

Each contig is represented as a node in the assembly graph and an edge between two nodes indicate their overlap (link) properties.
This step is to cluster the longest nodes (longer than $10$Kbp) into different sets, namely \emph{core} groups, based on the their degree (number of connections) and coverage values.
DBSCAN clustering algorithm \cite{Ester96adensity-based} is applied for this task by default.
The rationale is to approximate a coverage value of a significant contig (which can be splitted into more than 10,000 \emph{k-mers}) to be a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is a distance function based on Kullback-Leibner divergence \cite{Kullback1951information}, or relative entropy, of two Poisson distributions. Other than that, external binning tools can be employed for more complicated binning task, \EG{} metagenomics, and the results can be integrated into \npgraph{} for better performance. More details can be found on Supplementary Note \ref{supp:note1}.

\subsection*{Multiplicity estimation}
Due to the possible divergence of sequencing coverage relative to the real abundance of sequences, especially the short ones, an optimization step is carried out to alleviate this issue as shown in Supplementary Note \ref{supp:note2}. The coverage measures of nodes (which represent contigs) are spread throughout the graph via connecting edges for comparisons and calibrations prior to the multiplicity estimation. 
%The assignment of coverage to edges is also helpful to induce multiplicity of nodes in later step.

Based on the coverage values of all the edges and the graph's topology, we induce the copy numbers of every significant nodes (long contigs) in the final paths.
For each node, this could be done by investigating its adjacent edges and answering the questions of how many times it should be visited, from which abundance groups.
Multiplicities of insignificant nodes (of sequences with length less than $1,000$ bp) can be estimated in the same way but usually with less confident due to more complicated connections and greater variances of coverage values. 
For that reason, they are only used as augmented information to calculate candidate paths' score in the next step.

@@

We implement a modified stack-based version utilizing Dijkstra's shortest path finding algorithm~\cite{Dijkstra1959} to reduce the search space when using Depth-First Search (Supplementary Note~\ref{supp:note3}).
